---
output:
  rmarkdown::github_document
#  html_document: default
#  pdf_document: default
---
Red, Red Wine by Jason Horowitz
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.
setwd("/Users/jasonhorowitz/Desktop/DAND/t2-p2-project-EDA")
library(gridExtra)
library(plyr)
library(reshape2)
library(tidyverse) # includes ggplot2, dplyr, tidyr
library(GGally) # for scatterplot matrix
theme_set(theme_minimal(20))
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
rw <- read.csv("wineQualityReds.csv")
```
### Intro

Objective of this project is to perform EDA (exploratory data anlalysis) examining  relationships among variables in the red wine dataset. This is one of two datasets
presented by Udacity which relate to red and white variants of Portuguese Vinho 
Verde wine.

There are a total of 1599 observations (wines) in the dataset displaying twelve 
variables including eleven input variables (based on physicochemical tests) and 
 one output variable based on sensory data, a quality score.
 
The physico-chemical variables are (descriptive text included as indicated):

1. **fixed acidity** (tartaric acid - g / dm^3): most acids involved with wine are 
fixed or nonvolatile (do not evaporate readily).
2. **volatile acidity** (acetic acid - g / dm^3): the amount of acetic acid in wine, 
which at too high of levels can lead to an unpleasant, vinegar taste.
2. **citric acid** (g / dm^3): found in small quantities, citric acid can add 'freshness' 
and flavor to wines.
4. **residual sugar** (g / dm^3): the amount of sugar remaining after fermentation stops,
it's rare to find wines with less than 1 gram/liter and wines with greater than
45 grams/liter are considered sweet.
5. **chlorides** (sodium chloride - g / dm^3): the amount of salt in the wine
6. **free sulfur dioxide** (mg / dm^3): the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth 
and the oxidation of wine.
7. **total sulfur dioxide** (mg / dm^3): amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations
over 50 ppm, SO2 becomes evident in the nose and taste of wine.
8. **density** (g / cm^3): the density of water is close to that of water depending 
on the percent alcohol and sugar content.
9. **pH**: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 
14 (very basic); most wines are between 3-4 on the pH scale
10. **sulphates** (potassium sulphate - g / dm3): a wine additive which can contribute
to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant.
11. **alcohol** (% by volume): the percent alcohol content of the wine.
12. **quality** (score between 0 and 10):the median of at least 3 evaluations made by 
wine experts with grading  between 0 (very bad) and 10 (very excellent).

# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

```{r echo=FALSE, Univariate_Plots}
str(rw)

```

```{r}
# let's turn X into a factor
rw$X <- factor(rw$X)
# now let's rename X to something more appropriate like 'id'
colnames(rw)[1] <- "id"
# it may make sense to turn quality into a factor and then facet by quality
```

```{r}
summary(rw)
```

### Structure and Summary of Data

Our dataset consists of 12 variables (plus an index (X) which identifies the 
different wines). The variables are all number or integer data types. There are 
a total of nearly 1600 observations in the dataset. There are no missing values
and each row of the database equals one observation: so the dataset is both clean
and tidy.

### Histogram of Wine Quality Scores

For our first histogram, we plotted the wine quality scores. It is notable
that greater than 75% of the quality scores are concentrated in the middle (5-6) 
of the score range (1-10). The shape of the curve is normal-ish.

Also worth noting is that the Quality score is an integer data type despite the 
description that it is derived from observations of at least 3 experts. It would 
have been nice to have the precision of a decimal average for quality scores. On
the other hand, it simplifies things by making a bin = 1 optimal for displaying 
the distribution.

In fact, since the ratings are distributed across six integers (3-8), it's going
to make the most sense to convert this into a factor variable. The results will
coincide with the following ratings:

-  3 = "Awful"
-  4 = "Poor"
-  5 = "Fair"
-  6 = "Good"
-  7 = "Very Good"
-  8 = "Superior"
```{r}
# let's turn X into a factor
rw$quality <- factor(rw$quality)
str(rw$quality)
```

```{r}
ggplot(rw) +
  geom_histogram(aes(x = quality), color = "black",
                 fill = "#F79420", stat="count")
```

### Histograms of Other Variables

Let's look at histograms of the other variables adjusting the binwidth as appropriate
to have a look at the distribution of values. I will also perform transformations 
on plots that require normalization.

Hopefully, plotting the variables and some minor additional research into them will
give us ideas for further investigation in the bivariate and multivariate analysis 
to follow.

#### Fixed Acidity

```{r}
# fixed acidity
ggplot(rw) +
  geom_histogram(aes(x = fixed.acidity), color = "black",
                 fill = "#FFFF66", binwidth = .1)

```
Since this appears to have a long-tail distribution, we will transform the variable
using the log base 10 scale.
```{r}
ggplot(rw) +
  geom_histogram(aes(x = fixed.acidity), color = "black",
                 fill = "#FFFF66", binwidth = .02) +
  scale_x_log10() +
  xlab("fixed.acidity log10")
```
With this transformation, the distribution of the data follows a more normal distribution.

#### Volatile Acidity

```{r}
# volatile acidity
ggplot(rw) +
  geom_histogram(aes(x = volatile.acidity), color = "black",
                 fill = "#CCFFCC", binwidth = .01)
```
The distribution for volatile acidity also seems to be longtailed - and with some
outliers. Let's perform a similar transformation - also eliminating the top and
bottom 1% of the values.
```{r}
ggplot(rw) +
  geom_histogram(aes(x = volatile.acidity), color = "black",
                 fill = "#CCFFCC", binwidth = .02) +
  scale_x_log10() +
  xlim(quantile(rw$volatile.acidity, 0.01),quantile(rw$volatile.acidity, 0.99)) +
  xlab("volatile.acidity log10")
```
As with fixed acidity, the transformed volatile acidity data follows a somewhat more
normal distribution.

#### Citric Acid

```{r}
ggplot(rw) +
  geom_histogram(aes(x = citric.acid), color = "black",
                 fill = "#FF66CC", binwidth = .02)
```
A signicant number of wines have citric acid near 0. Let's check how many actually 
have no trace of citric acid.
```{r}
# r code to count number of items at a specific value
no_citric <- subset(rw, citric.acid == 0)
dim(no_citric)
dim(no_citric)[1]/dim(rw)[1]
```
So 132 items (8.2% of the observations) contain no trace of citric acid. According
to x of wine this makes sense because XXXXX

#### Residual Sugar
```{r}
ggplot(rw) +
  geom_histogram(aes(x = residual.sugar), color = "black",
                 fill = "#993300", binwidth = .05)
```
Residual sugar seems to have a very long tail, so let's perform a log 10 transformation.

```{r}
ggplot(rw) +
  geom_histogram(aes(x = residual.sugar), color = "black",
                 fill = "#993300", binwidth = .05) +
  scale_x_log10() +
  xlab("residual.sugar_log10")
  
```
Still a long tail, but the middle of the distribution is more clearly normalish. Let's try doing this a different way to see if we can eliminate the outliers.
```{r}
ggplot(rw) +
  geom_histogram(aes(x = residual.sugar), color = "black",
                 fill = "#993300", binwidth = .1) +
  scale_x_continuous(lim = c(0, quantile(rw$residual.sugar, 0.95))) +
  xlab("residual.sugar_95th percentiles")
```
That's a little better.

#### Chlorides

```{r}
ggplot(rw) +
  geom_histogram(aes(x = chlorides), color = "black",
                 fill = "#9999FF", binwidth = .005)
```
Chlorides definitely has a long-tail, so let's try removing the outliers.

```{r}
ggplot(rw) +
  geom_histogram(aes(x = chlorides), color = "black",
                     fill = "#9999FF", binwidth = .002) +
  scale_x_continuous(lim = c(quantile(rw$chlorides, 0.01), quantile(rw$chlorides, 0.95))) +
  xlab('chlorides_95thPercentile')

```
This plot definitely is more normal in its appearance than the previous one.

#### Sulphates

```{r}
ggplot(rw) +
  geom_histogram(aes(x = sulphates), color = "black",
                 fill = "#CCFF00", binwidth = .05)

```

#### pH (Overall Acidity)
```{r}
ggplot(rw) +
  geom_histogram(aes(x = pH), color = "black",
                 fill = "#CC0033", binwidth = .01)

```

#### Density

```{r}
ggplot(rw) +
  geom_histogram(aes(x = density), color = "black",
                 fill = "#660033", binwidth = .0001)
```

#### Total Sulfur Dioxide
```{r}
ggplot(rw) +
  geom_histogram(aes(x = total.sulfur.dioxide), color = "black",
                 fill = "#669933", binwidth = 1)
```

#### Free Sulfur Dioxide

```{r}
ggplot(rw) +
  geom_histogram(aes(x = free.sulfur.dioxide), color = "black",
                 fill = "#99FF99", binwidth = 1)
```


#### Alcohol Content % By Volume

```{r}
# alcohol content % by volume
ggplot(rw) +
  geom_histogram(aes(x = alcohol), color = "black",
                 fill = "#FF6666", binwidth = .1)
```

> **Tip**: Make sure that you leave a blank line between the start / end of
each code block and the end / start of your Markdown text so that it is
formatted nicely in the knitted text. Note as well that text on consecutive
lines is treated as a single space. Make sure you have a blank line between
your paragraphs so that they too are formatted for easy readability.

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?
missing measurements - how to accomplish in R
### What is/are the main feature(s) of interest in your dataset?
Biggest question for me is how/whether the different input variables correlate 
or predict the output variable, quality. Am particularly curious to see how
alchohol level correlates with quality
Will also be interesting to see how the different acidity-related variables 
correlate with each other and with qulaity.
### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?
I did not create any new variables, but I did convert two of the existing variables:
X (renaming it 'ID') and quality into factor variables. Converting them to factors
means that R won't try to summarize them the same way as it does integers.
### Of the features you investigated, were there any unusual distributions? \
## Did you perform any operations on the data to tidy, adjust, or change the form \
### of the data? If so, why did you do this?
This is a relatively small dataset so log or exponential transformations don't
do anything to make histograms easier to understand. The only thing I did was to 
adjust the binwidth to make the counts at different values easier to read.


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}
# scatterplot matrix
rw_subset = rw[, c('volatile.acidity', 'residual.sugar', 'density', 'pH', 'alcohol', 'quality')]
head(rw_subset,3)
ggpairs(rw_subset[sample.int(nrow(rw_subset),1000), ])
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!